OpenMPI implementation of Burrows-Wheeler Transform for substring matching applied to genomics

### 2023-08-27 report/paper.tex
.

\section{Biological context and Data}
\label{Biological context and Data}
The algorithm considered in this project is being used to find a gene within the genome of an organism. The genome is made up of DNA and contains the set of genes of an organism which are responsible for its traits and characteristics.
DNA is a long double helix molecule composed by a sequence of nucleotides\cite{nucleotide}, each nucleotide is composed by deoxyribose sugar, a phosphate group and a nucleobase. There are four types of nucleobases in DNA: guanine (G), adenine (A), cytosine (C) and thymine (T).
In bioinformatics applications, the genome and the genes are usually considered as long strings of computer code composed by four possible letters (G,A,C,T). \\
The genomes used for this analysis have been taken from the NCBI website\cite{ncbi}. \\
Genomes of different sizes have been used in order to study how the performance changes with respect to the input dataset dimension:
\begin{itemize}
    \item \emph{Escherichia coli}, bacteria commonly found in the lower intestine of warm-blooded organisms\cite{ecoli}, 4.6MB \cite{ecoli dataset}
    \item \emph{Entamoeba invadens}, an amoebozoa parasite of reptiles\cite{entamoeba}, 40.9MB \cite{entamoeba dataset}
    \item \emph{Formica exsecta}, is a species of ant found from Western Europe to Asia \cite{formica}, 277.6MB \cite{formica dataset}
\end{itemize}

@@

\subsection{Analysis of the serial algorithm}
In this section, it will be discussed how the serial version of the code has been implemented and the computational costs associated to the different steps of the algorithm, in terms of number of instructions and in relation to the input size. \\
Before applying the previously described BWT algorithm it is necessary to read and preprocess both the genome and the gene file. The path of these files are passed to the \emph{main()} function as parameters when the program is executed. The preprocessing involves removing spaces, new-lines, characters different from the nucleobases symbols, converting lower case letters to upper case and adding a termination character at the end of the preprocessed strings; all these operations are necessary for the correct execution of the algorithm. The computational costs of both the reading and preprocessing are linear with the size of the input file: 
$$ O(genomeSize + patternSize) $$
The Burrows-Wheeler Transform and the pattern matching are implemented in three phases: \\
\begin{enumerate}
    \item \textbf{Suffix Array computation}: as it can be noticed by looking at the code implementation, it contains two for loops iterating over the genome string and an execution of the quicksort algorithm, so the computational cost of this phase can be computed as $$O(2*genomeSize + genomeSize*ln(genomeSize))$$
    \item \textbf{Last column computation}: since it contains basically a for loop iteration over the genome string, its computational cost can be computed as $$O(genomeSize)$$
    \item \textbf{Pattern search}: it iterates the genome string 4 times in order to count the number of occurences for each of the four nucleobases; then assuming we are in the case that a match is found we have to iterate over all the nucleobases in the pattern and for each iteration we have to execute the rank function twice; the cost of the rank computation depends on the index from which we want to count the number of occurences of a character in the BWT array, so on average we can say it is $genomeSize/2$. So we can estimate the total computational cost of the pattern search as
    $$O(4*genomeSize + patternSize*genomeSize)$$
\end{enumerate}
The following table summarizes the computational costs of the different steps. More precise results could be estimated using profiling tools.
\begin{center}
\begin{tabular}{|| c c ||}
\hline
Step & Cost (N. operations) \\
\hline
 Read data & $O(genSize + patSize)$ \\ 
 Preprocess data & $O(genSize + patSize)$ \\  
 Suffix array & $O(2*genSize + genSize*ln(genSize))$ \\
 BWT last column & $O(genSize)$ \\
 Pattern search & $O(4*genSize + patSize*genSize)$ \\
\hline
\end{tabular}
\end{center}

\section{A priori study of available parallelism}
The serial version of the code has been analyzed in order to understand which parts can be parallelized and what enhancement can be theoretically expected by executing the code on multiple CPUs cooperating with each other in parallel.\\
In order to estimate the improvement in performance, in terms of speedup, the \textbf{Amdalh's law} can be used:
$$ Speedup(N) = \frac{1}{S + \frac{P}{N}} $$
Where N is the number of nodes (virtual CPUs) that are exploited, S is the fraction of the code that cannot be parallelized and P is the fraction that can be parallelized. It can be noticed that obviously S+P=1 and the greater is P with respect to S the more speedup is obtained, at least theoretically.\\
In the following analysis parts of the code execution that are less relevant, such as the import of the libraries and the check of the parameters passed to the \emph{main()} function, have been neglected and only the most important blocks, discussed in the previous section, have been considered.\\
The only parts that cannot be parallelized, at least not efficiently or simply, are the reading and preprocessing of the genome and gene files. Even the string matching algorithm based on the Burrows-Wheeler Transform is intrinsically serial, due to loop-carried dependencies, but it is possible to parallelize it by distributing chunks of the genome on different nodes so that the nodes can perform the BWT pattern search in parallel each on a different genome portion, smaller than the whole dataset. The choice of this strategy is also motivated by the fact that, as it can be noticed from the table in the previous section, the computational costs of the operations are highly dependent on the genome dataset size, so we can expect an enhancement by distributing it among different nodes. The methods and issues regarding the parallel implementation will be discussed in greater details in the next section. Whereas it is not possible to distribute the pattern string across different nodes because they all need the whole pattern to be able to find it.\\

>>>>

\section{Biological context and Data}
\label{Biological context and Data}
The algorithm considered in this project is being used to find a gene within the genome of an organism. The genome is made up of DNA and contains the set of genes of an organism which are responsible for its traits and characteristics.
DNA is a long double helix molecule composed by a sequence of nucleotides\cite{nucleotide}, each nucleotide is composed by deoxyribose sugar, a phosphate group and a nucleobase. There are four types of nucleobases in DNA: guanine (G), adenine (A), cytosine (C) and thymine (T).
In bioinformatics applications, the genome and the genes are usually considered as long strings of computer code composed by four possible letters (G,A,C,T). \\
The genomes used for this analysis have been taken from the NCBI website\cite{ncbi}. \\
Genomes of different sizes have been used in order to study how the performance changes with respect to the input dataset dimension:
\begin{itemize}
    \item \emph{Escherichia Coli}, bacteria commonly found in the lower intestine of warm-blooded organisms\cite{ecoli}, 4.6MB \cite{ecoli dataset}
    \item \emph{Entamoeba Invadens}, an amoebozoa parasite of reptiles\cite{entamoeba}, 40.9MB \cite{entamoeba dataset}
    \item \emph{Formica Exsecta}, is a species of ant found from Western Europe to Asia \cite{formica}, 277.6MB \cite{formica dataset}
\end{itemize}

@@

\subsection{Analysis of the serial algorithm}
In this section, it will be discussed how the serial version of the code has been implemented and the computational costs associated to the different steps of the algorithm, in terms of number of instructions and in relation to the input size. \\
Before applying the previously described BWT algorithm it is necessary to read and preprocess both the genome and the gene file. The path of these files are passed to the \emph{main()} function as parameters when the program is executed. The preprocessing involves removing spaces, new-lines, characters different from the nucleobases symbols, converting lower case letters to upper case and adding a termination character at the end of the preprocessed strings; all these operations are necessary for the correct execution of the algorithm. The computational costs of both the reading and preprocessing are linear with the size of the input file: 
$$ O(genomeSize + patternSize) $$
The Burrows-Wheeler Transform and the pattern matching are implemented in three phases: \\
\begin{enumerate}
    \item \textbf{Suffix Array computation}: as it can be noticed by looking at the code implementation, it contains two for loops iterating over the genome string and an execution of the quicksort algorithm, so the computational cost of this phase can be computed as $$O(2*genomeSize + genomeSize*ln(genomeSize))$$
    \item \textbf{Last column computation}: since it contains basically a for loop iteration over the genome string, its computational cost can be computed as $$O(genomeSize)$$
    \item \textbf{Pattern search}: it iterates the genome string 4 times in order to count the number of occurences for each of the four nucleobases; then assuming we are in the case that a match is found we have to iterate over all the nucleobases in the pattern and for each iteration we have to execute the rank function twice; the cost of the rank computation depends on the index from which we want to count the number of occurences of a character in the BWT array, so on average we can say it is $genomeSize/2$. So we can estimate the total computational cost of the pattern search as
    $$O(4*genomeSize + patternSize*genomeSize)$$
\end{enumerate}
The following table summarizes the computational costs of the different steps.
\begin{center}
\begin{tabular}{|| c c ||}
\hline
Step & Cost (N. operations) \\
\hline
 Read data & $O(genSize + patSize)$ \\ 
 Preprocess data & $O(genSize + patSize)$ \\  
 Suffix array & $O(2*genSize + genSize*ln(genSize))$ \\
 BWT last column & $O(genSize)$ \\
 Pattern search & $O(4*genSize + patSize*genSize)$ \\
\hline
\end{tabular}
\end{center}
In order to have a more realistic performance evaluation of each section of the code, execution times have been measured considering the combinations of genome and pattern sizes shown in the following table (the same ones will be used for testing the parallel implementation on GCP).
\begin{center}
\begin{tabular}{|| c c c ||}
\hline
Dataset & Genome size & Pattern size \\
\hline
 Escherichia Coli & 4699745 & 933 \\
 Entamoeba Invadens & 41522539 & 654 \\
 Formica Exsecta & 283043175 & 200 \\
\hline
\end{tabular}
\end{center}
The average times measured are shown in the table below. The serial code has been executed on a MacBook Pro, Apple M1 Pro chip, 16GB RAM. More precise results could be obtained by using more advanced profiling tools and testing the code on different devices and with more input combinations.
\begin{center}
\begin{tabular}{|| c c c c ||}
\hline
Dataset & Step & Time(s) & \% \\
\hline
 E. Coli & Read data & 0,105 & 0,629 \\
    & Preprocess data & 0,086 & 0,516 \\
    & Suffix array & 1,669 & 10,006 \\
    & BWT last column & 0,012 & 0,072 \\
    & Pattern search & 14,808 & 88,777 \\
    & Total & 16,68 & 100 \\
\hline
E. Invadens & Read data & 0,712 & 0,681 \\
    & Preprocess data & 0,742 & 0,71 \\
    & Suffix array & 19,903 & 19,035 \\
    & BWT last column & 0,403 & 0,385 \\
    & Pattern search & 82,801 & 79,189 \\
    & Total & 104,561 & 100 \\
\hline
F. Exsecta & Read data & 4,738 & 1,581 \\
    & Preprocess data & 4,864 & 1,623 \\
    & Suffix array & 152,034 & 50,732 \\
    & BWT last column & 3,357 & 1,12 \\
    & Pattern search & 134,689 & 44,944 \\
    & Total & 299,682 & 100 \\
\hline
\end{tabular}
\label{fig:times_table}
\end{center}
This results are in line with what can be expected from the theoretical cost analysis previously done.

\section{A priori study of available parallelism}
The serial version of the code has been analyzed in order to understand which parts can be parallelized and what enhancement can be theoretically expected by executing the code on multiple CPUs cooperating with each other in parallel.\\
In order to estimate the improvement in performance, in terms of speedup, the \textbf{Amdalh's law} can be used:
$$ Speedup(N) = \frac{1}{S + \frac{P}{N}} $$
Where N is the number of nodes (virtual CPUs) that are exploited, S is the fraction of the code that cannot be parallelized and P is the fraction that can be parallelized. It can be noticed that obviously S+P=1 and the greater is P with respect to S the more speedup is obtained, at least theoretically.\\
The only parts that cannot be parallelized, at least not efficiently or simply, are the reading and preprocessing of the genome and gene files. Even the string matching algorithm based on the Burrows-Wheeler Transform is intrinsically serial, due to loop-carried dependencies, but it is possible to parallelize it by distributing chunks of the genome on different nodes so that the nodes can perform the BWT pattern search in parallel each on a different genome portion, smaller than the whole dataset. The choice of this strategy is also motivated by the fact that, as it can be noticed from the table in the previous section, the computational costs of the operations are highly dependent on the genome dataset size, so we can expect an enhancement by distributing it among different nodes. Whereas it is not possible to distribute the pattern string across different nodes because they all need the whole pattern to be able to find it. The methods and issues regarding the parallel implementation will be discussed in greater details in the next section. \\
The Amdalh's law has been applied after computing the S and P values for each dataset using the execution times in the table \ref{fig:times_table}. The expected theoretical speedups calculated for all datasets are shown in Figure \ref{fig:theoretical-speedup-label}.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{images/theoretical_speedup.png}
    \caption{Theoretical Speedup}
    \label{fig:theoretical-speedup-label}
\end{figure}
\\It can be noticed that for larger datasets the expected speedup is lower than for smaller ones due to the non-parallelizable fraction S being greater (as shown in the execution times table \ref{fig:times_table}) probably because there are overheads due to the allocation and management of a bigger amount of memory.
